Three Algorithms for Merging Navigable Graphs

### 2025-04-23 paper/main.tex
fix pseudocode

We provide 3 merge algorithms: "Naive Merge", "Merge1", and "Merge2". We also refer to these algorithm as layer-merge algorithms. "Naive Merge" is a straight forward method, while "Merge1" and "Merge2" are more advanced. Together, these two algorithms constitute the main contribution of the work. The HNSW structure is a sequence of graphs, called layers. We start from description of a general algorithm for merging multi layers structure.

\subsection{HNSW General Merge}

HNSW-General-Merge algorithm \ref{alg:general_merge} is a straight forward. It uses one of the layer-merge algorithm to merge each HNSW layer. To able layer-merge algorithm to perform fast search using multi-layer structure of HNSW, HNSW-General-Merge algorithm passes to layer merge algorithm both HNSW structures $\mathcal{H}_a, \mathcal{H}_b$, and the layer number $\ell=i$ to which the merge should be performed. An order of merging layers doesn't mater. Thus merging of layers can be done in parallel.

\begin{algorithm}
\caption{\textsc{HNSW-General-Merge}($\mathcal{H}_a, \mathcal{H}_b$)}\label{alg:general_merge}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$ \\
\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
\begin{algorithmic}[1]

\For{$i = 0 \textbf{ to } \max(|\mathcal{H}_a|, |\mathcal{H}_b|) $}
    \State $G^c_i \gets \text{Merge}(\mathcal{H}_a, \mathcal{H}_b, \ell=i, \text{algorithm specific params})$
\EndFor

\State $\mathcal{H}_c \gets (G^c_i)_i^{\max(|\mathcal{H}_a|, |\mathcal{H}_b|)}$

\State \Return $\mathcal{H}^c$
\end{algorithmic}
\end{algorithm}

\subsection{Naive Merge}
The Naive Merge algorithm \ref{alg:merge_naive} surprisingly is very simple. It takes one vertex $v^*$ by one from the graph $G^a$. Then algorithm prepares the candidate set $\mathcal{C}$ for a new neighborhood (lines 6,7). It uses \textsc{HNSW-Search} to get $m$-closest vertices to $v^*$ in the graph $G^b$, and combines with an old neighborhood of the vertex $v^*$. After that in the line 8 the candidate set $\mathcal{C}$ is passed into the neighborhood construction function, which selects the vertices for the new merged neighborhood for $v^*$.
Lines 9-12 contains the same logic for the forming new neighborhoods of graph $G^b$

\begin{algorithm}
\caption{\textsc{Merge-Naive}($\mathcal{H}_a, \mathcal{H}_b,  \ell, \text{search\_ef}$)}\label{alg:merge_naive}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i)$, $\mathcal{H}_b = (G^b_i)$, and the level number $\ell$, search parameter $\text{search\_ef}$ \\
\textbf{Ensure:} Merged graph $G^c$ 
\begin{algorithmic}[1]
\State $V^c \gets V^a \cup V^b$
\State $E^c \gets \emptyset$
\State $G^a \gets G^a_{\ell}$
\State $G^b \gets G^b_{\ell}$
% \State $m \gets \mathcal{H}_a.m_0$ if $\ell = 0$ else $\mathcal{H}_a.m$

\For{$v^* \in V^a$}
    \State $\mathcal{C}^b \gets \textsc{HNSW-Search}(\mathcal{H} = \mathcal{H}^b, q = v^*, v_0, k = m, L = \text{search\_ef}, \ell)$
    \State $\mathcal{C} \gets \{v : (v^*, v) \in E^a \} \cup \mathcal{C}^b$
    \State $E^c \gets E^c \cup \{(v^*, v) : v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$
\EndFor

\For{$v^* \in V^b$}
    \State $\mathcal{C}^a \gets \textsc{HNSW-Search}(\mathcal{H} = \mathcal{H}^a, q = v^*, v_0, k = m, L = \text{search\_ef}, \ell)$
    \State $\mathcal{C} \gets \{v : (v^*, v) \in E^b \} \cup \mathcal{C}^a$
    \State $E^c \gets E^c \cup \{(v^*, v) : v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$
\EndFor

@@

@@

\begin{algorithm}
\caption{\textsc{Merge1}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge1}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the merging layer number $\ell$, the size of the forming neighborhoods $m \in \mathbb{N}$, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k} \in \mathbb{N}$ \\
\textbf{Ensure:}  Merged graph $G^c$ 
\begin{algorithmic}[1]
\State $V^c \gets V^a \cup V^b$
\State $E^c \gets \emptyset$
\State $G^a \gets G^a_{\ell} $
\State $G^b \gets G^b_{\ell} $
\State $\mathcal{V}_{not\_done} \gets V^a$

@@

\begin{algorithm}
\caption{\textsc{Merge2}($\mathcal{H}^a, \mathcal{H}^b,  \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge2}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the number of merged layers, the size of the forming neighborhoods $m \in \mathbb{N} $, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M \in \mathbb{N}$ \\
\textbf{Ensure:}  Merged graph $G^c$ 
\begin{algorithmic}[1]

\State $V^c \gets V^a \cup V^b$
\State $E^c \gets \emptyset$
\State $G^a \gets G^a_{\ell} $
\State $G^b \gets G^b_{\ell} $
% \State $m \gets \mathcal{H}_a.m_0 \text{ if } \ell = 0 \text{ else } \mathcal{H}_a.m$
\State $\mathcal{V}_{not\_done} \gets V^a \cup V^b$

>>>>

%We provide 3 merge algorithms: "Naive Merge", "Merge1", and "Merge2". We also refer to these algorithm as layer-merge algorithms. "Naive Merge" is a straight forward method, while "Merge1" and "Merge2" are more advanced. Together, these two algorithms constitute the main contribution of the work. The HNSW structure is a sequence of graphs, called layers. We start from description of a general algorithm for merging multi layers structure.
We now present three algorithms for merging HNSW graphs: \textsc{Merge-Naive}, \textsc{Merge1}, and \textsc{Merge2}. These algorithms operate layer by layer. We first describe the general procedure for merging the multi-layer HNSW structure, followed by the specific logic for merging a single layer (\textit{layer-merge algorithms}).

\subsection{HNSW General Merge Framework}

The \textsc{HNSW-General-Merge} algorithm (Algorithm~\ref{alg:general_merge}) provides a framework for merging two HNSW structures, $\mathcal{H}_a$ and $\mathcal{H}_b$. It iterates through each layer level and applies a chosen layer-merge algorithm (\textsc{Merge-Naive}, \textsc{Merge1}, or \textsc{Merge2}) to combine the corresponding layers $G^a_i$ and $G^b_i$. The layer-merge algorithms require access to the full HNSW structures $\mathcal{H}_a, \mathcal{H}_b$ to perform efficient searches (using \textsc{HNSW-Search} or \textsc{LocalSearch}) down to the target layer $\ell=i$.

Let $l_{\max}^a$ and $l_{\max}^b$ be the maximum layer indices for $\mathcal{H}_a$ and $\mathcal{H}_b$, respectively. The merged graph $\mathcal{H}_c$ will have $l_{\max}^c = \max(l_{\max}^a, l_{\max}^b)$ layers.

\begin{algorithm}
\caption{\textsc{HNSW-General-Merge}($\mathcal{H}_a, \mathcal{H}_b, \text{LayerMergeAlgo}, \text{params}$)}\label{alg:general_merge}
\textbf{Input:} HNSW graphs $\mathcal{H}_a$, $\mathcal{H}_b $. Chosen layer merge algorithm $\text{LayerMergeAlgo}$ (e.g., \textsc{Merge-Naive}). Algorithm-specific parameters $\text{params}$. \\
\textbf{Output:} Merged HNSW graph $\mathcal{H}_c = (G^c_i)_{i=0}^{l_{\max}^c}$
\begin{algorithmic}[1]
\State $l_{\max}^a \gets \mathcal{H}_a.\text{getMaxLayerNumber()}$
\State $l_{\max}^b \gets \mathcal{H}_b.\text{getMaxLayerNumber()}$
\State $l_{\max}^c \gets \max(l_{\max}^a, l_{\max}^b)$
\For{$i = 0 \textbf{ to } l_{\max}^c$} \Comment{Merge each layer}    
    \State $G^c_i \gets \text{LayerMergeAlgo}(\mathcal{H}_a, \mathcal{H}_b, \ell=i, \text{params})$ \Comment{Calls e.g., Merge-Naive}
\EndFor
\State $\mathcal{H}_c \gets (G^c_i)_{i=0}^{l_{\max}^c}$ % Construction is implicit in the loop

\State \Return $\mathcal{H}_c$
\end{algorithmic}
\end{algorithm}

Note that the merging of different layers (iterations of the loop in Algorithm~\ref{alg:general_merge}) is independent and can potentially be parallelized.

%\subsection{Naive Merge}
%The Naive Merge algorithm \ref{alg:merge_naive} surprisingly is very simple. It takes one vertex $v^*$ by one from the graph $G^a$. Then algorithm prepares the candidate set $\mathcal{C}$ for a new neighborhood (lines 6,7). It uses \textsc{HNSW-Search} to get $m$-closest vertices to $v^*$ in the graph $G^b$, and combines with an old neighborhood of the vertex $v^*$. After that in the line 8 the candidate set $\mathcal{C}$ is passed into the neighborhood construction function, which selects the vertices for the new merged neighborhood for $v^*$.
%Lines 9-12 contains the same logic for the forming new neighborhoods of graph $G^b$

\subsection{Naive Merge}
The \textsc{Merge-Naive} algorithm (Algorithm~\ref{alg:merge_naive}) provides a straightforward method for merging a single layer $\ell$. For each vertex $v^*$ originating from graph $G^a_\ell$, it searches for its potential neighbors in graph $G^b_\ell$ using the full \textsc{HNSW-Search} procedure down to layer $\ell$. These candidates are combined with the original neighbors of $v^*$ in $G^a_\ell$. The final neighborhood for $v^*$ in the merged graph $G^c_\ell$ is selected from this combined set using the chosen \textsc{NeighborhoodConstruction} strategy. The process is repeated for vertices originating from $G^b_\ell$.

\begin{algorithm}
\caption{\textsc{Merge-Naive}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{NeighborhoodConstruction}, m, \text{search\_ef}, v_{entry}$)}\label{alg:merge_naive}
\textbf{Input:} HNSW graphs $\mathcal{H}_a$, $\mathcal{H}_b$; target layer $\ell$; Neighborhood construction function \textsc{NeighborhoodConstruction}; target neighborhood size $m$; search parameter $\text{search\_ef}$; entry point $v_{entry}$ (e.g., from $\mathcal{H}_a$ or $\mathcal{H}_b$) \\
\textbf{Output:} Merged graph $G^c$
\begin{algorithmic}[1]
\State $G^a \gets \mathcal{H}_a\text{.GetLayer}(\ell) $; $G^b \gets\mathcal{H}_a\text{.GetLayer}(\ell)$ 
\State $V^a \gets \text{vertices}(G^a)$; $V^b \gets \text{vertices}(G^b)$
\State $E^a \gets \text{edges}(G^a)$; $E^b \gets \text{edges}(G^b)$
\State $V^c \gets V^a \cup V^b$; $E^c \gets \emptyset$ \Comment{Initialize vertices and edges for the merged graph}
% \State $m \gets \dots$ % Now passed as input parameter

\For{$v^* \in V^a$} \Comment{Process vertices from graph A}
    \State $\mathcal{C}^b \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}_b, q=v^*, v_{entry}=v_{entry}^b, k=m, L=\text{search\_ef}, \ell=\ell)$ \Comment{Find candidates in B}
    \State $\mathcal{C} \gets \{v \mid (v^*, v) \in E^a \} \cup \mathcal{C}^b$  \Comment{Combine original neighbors and candidates from B}
    \State $E^c \gets E^c \cup \{(v^*, v) | v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$\Comment{Add new neighborhood to set of edges}
\EndFor

\For{$v^* \in V^b$} \Comment{Process vertices from graph B}
    \State $\mathcal{C}^a \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}_a, q=v^*, v_{entry}=v_{entry}^a, k=m, L=\text{search\_ef}, \ell_{target}=\ell)$ \Comment{Find candidates in A}
    \State $\mathcal{C} \gets \{v \mid (v^*, v) \in E^b \} \cup \mathcal{C}^a$  \Comment{Combine original neighbors and candidates from A}
    \State $E^c \gets E^c \cup \{(v^*, v) | v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$\Comment{Add new neighborhood to set of edges}
\EndFor

@@

This naive approach is simple but computationally intensive due to repeated \textsc{HNSW-Search} calls, each potentially traversing multiple layers from the top down.

@@

\begin{algorithm}
\caption{\textsc{Merge1}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge1}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the merging layer number $\ell$, the size of the forming neighborhoods $m \in \mathbb{N}$, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k} \in \mathbb{N}$ \\
\textbf{Ensure:}  Merged graph $G^c$ 
\begin{algorithmic}[1]
\State $G^a \gets \mathcal{H}_a\text{.GetLayer}(\ell) $; $G^b \gets\mathcal{H}_a\text{.GetLayer}(\ell)$ 
\State $V^a \gets \text{vertices}(G^a)$; $V^b \gets \text{vertices}(G^b)$
\State $E^a \gets \text{edges}(G^a)$; $E^b \gets \text{edges}(G^b)$
\State $V^c \gets V^a \cup V^b$; $E^c \gets \emptyset$ 
\State $\mathcal{V}_{not\_done} \gets V^a$

@@

\begin{algorithm}
    \caption{\textsc{Merge2}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge2}
    \textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i)$, $\mathcal{H}_b = (G^b_i)$, the merging layer $\ell$, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k} \in \mathbb{N}$, neighborhood sizes $M, m \in \mathbb{N}$ \\
    \textbf{Ensure:} Merged graph $G^c$ 
\begin{algorithmic}[1]

\State $G^a \gets \mathcal{H}_a\text{.GetLayer}(\ell) $; $G^b \gets\mathcal{H}_a\text{.GetLayer}(\ell)$ 
\State $V^a \gets \text{vertices}(G^a)$; $V^b \gets \text{vertices}(G^b)$
\State $E^a \gets \text{edges}(G^a)$; $E^b \gets \text{edges}(G^b)$
\State $V^c \gets V^a \cup V^b$; $E^c \gets \emptyset$
% \State $m \gets \mathcal{H}_a.m_0 \text{ if } \ell = 0 \text{ else } \mathcal{H}_a.m$
\State $\mathcal{V}_{not\_done} \gets V^a \cup V^b$
